### 2013-06-30 doc/dissertation/document/backmatter/diffusion_problem.tex
cleaning up first sections of parallel chapter

\chapter{Two-Dimensional One-Speed Neutron Diffusion Model Problem}
\label{chap:diffusion_problem}
For many numerical experiments in this work, we choose the one-speed,
two-dimensional neutron diffusion equation as a model transport
problem \citep{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogeneous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

>>>>

\chapter{Two-Dimensional One-Speed Neutron Diffusion Model Problem}
\label{chap:diffusion_problem}
For many numerical experiments in this work, we choose the one-speed,
two-dimensional neutron diffusion equation as a model neutron
transport problem \citep{duderstadt_nuclear_1976}:
\begin{equation}
  -\boldsymbol{\nabla} \cdot D \boldsymbol{\nabla} \phi + \Sigma_a
  \phi = S\:,
  \label{eq:diffusion_eq}
\end{equation}
where $\phi$ is the neutron flux, $\Sigma_a$ is the absorption cross
section, and $S$ is the source of neutrons. In addition, $D$ is the
diffusion coefficient defined as:
\begin{equation}
  D = \frac{1}{3 ( \Sigma_t - \bar{\mu}\Sigma_s )}\:,
  \label{eq:diffusion_coeff}
\end{equation}
where $\Sigma_s$ is the scattering cross section, $\Sigma_t = \Sigma_a
+ \Sigma_s$ is the total cross section, and $\bar{\mu}$ is the cosine
of the average scattering angle. For simplicity, we will take
$\bar{\mu} = 0$ for our analysis giving $D=(3 \Sigma_t)^{-1}$. In
addition, to further simplify we will assume a homogeneous domain such
that the cross sections remain constant throughout. Doing this permits
us to rewrite Eq~(\ref{eq:diffusion_eq}) as:
\begin{equation}
  -D \boldsymbol{\nabla}^2 \phi + \Sigma_a \phi = S\:.
  \label{eq:diffusion_eq_simple}
\end{equation}

### 2013-06-30 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
finished adding replication data for titan
cleaning up first sections of parallel chapter

\chapter{Parallel Monte Carlo Synthetic Acceleration}
\label{ch:parallel_methods}
For MCSA to be viable at the production scale, scalable parallel
implementations of the algorithm are required. Reviewing the
literature, MCSA has yet to be parallelized and the Neumann-Ulam Monte
Carlo method on which it is built has only been parallelized through
history-level parallelism with full domain replication. In order to
solve large linear systems with MCSA, a domain-decomposed parallel
strategy is required. In the formulation of a parallel MCSA algorithm,
we recognize that the algorithm occurs in two stages, an outer
iteration performing fixed point iteration and applying the
correction, and an inner Monte Carlo solver that is generating the
correction via the adjoint or forward methods. The parallel aspects of
both these components must be considered. Therefore, we will develop
and implement parallel algorithms for the Neumann-Ulam and MCSA
methods leveraging both the knowledge gained from the general parallel
implementations of Krylov methods reviewed in
Appendix~\ref{ch:linear_problem} and modern parallel strategies for
domain decomposed Monte Carlo as developed by the reactor physics
community.

In this chapter we briefly review particle transport methods for
domain decomposed Monte Carlo. Considering a domain decomposed
Neumann-Ulam method, we derive an analytic framework based on
algebraic quantities from which to estimate its performance when
applied to linear systems. We then devise a parallel algorithm for the
Neumann-Ulam method based on the multiple-set overlapping-domain
decomposition algorithm and a parallel algorithm for the MCSA that
leverages the parallel Monte Carlo algorithm and general parallel
matrix-vector operations. In identical fashion to the serial
computations presented in Chapter~\ref{ch:spn_equations}, we use
parallel solutions for the $SP_N$ equations for the fuel assembly
criticality problem to verify the correctness of the parallel
algorithm and implementation by comparing the numerical results
against production Krylov methods. Using the verified implementation
of the algorithm, we perform parallel scaling studings to test its
performance on both a small computing cluster and a leadership class
machine and compare this performance to the production Krylov methods.

@@

To motivate this problem, consider the square domain presented in
Figure~\ref{fig:ddmc_example}. Here, we imagine a Monte Carlo particle
transport problem. If the domain were decomposed into 9 pieces as
shown by the domain, each of those pieces could be parallel process in
the computation. In Figure~\ref{fig:ddmc_example}, the tracks of 3
particles that are born in the center domain are shown. For example,
particle A is first transported from the center domain to the domain
directly to the left. Before the scattering event in the new domain
may be processed, particle A must be communicated between between the
two parallel processes that own those domains. For any given particle
in the transport simulation, this communication event may hardly occur
during the lifetime of the particle (particle B), or may occur many
times (particle C) depending on the problem parameters. Scalable
parallel algorithms for domain Monte Carlo are those that handle this
domain-to-domain communication of particles effectively and balance it
with the physics computations.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/ddmc_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Domain decomposed Monte Carlo transport example
      illustrating how domain decomposition requires parallel
      communication of particles.}}
  \label{fig:ddmc_example}
\end{figure}

For a domain decomposed Neumann-Ulam method, the situation is very
much the same where instead the transport process is now discrete and
the "physics" of the transport process is the Monte Carlo game with
probabilities and weights described by
Eq~(\ref{eq:neumann_ulam_decomposition}) for the given linear
problem. Figure~\ref{fig:ddnu_example} gives the domain decomposed
Neumann-Ulam analog of particle transport problem in
Figure~\ref{fig:ddmc_example}. Imagine in this problem that we are
solving for the monoenergetic scalar neutron flux as in the model
diffusion problem presented in Appendix~\ref{chap:diffusion_problem}
such that the solution has only a spatial dependence. In
Figure~\ref{fig:ddnu_example}, the domain has now been discretized by
a mesh and again decomposed into 9 pieces. As we play the Monte Carlo
game presented in Chapter~\ref{ch:stochastic_methods}, the
discretization and resulting Neumann-Ulam decomposition of the problem
describes how each state in the system is coupled to the other states
in the system through the set of equations that forms the linear
system. As discrete states (or rows in the linear system) are reached
by stochastic histories that do not exist in the local domain, the
histories must be computed in an analogous fashion to the particles in
the previous example. Again consider history A in
Figure~\ref{fig:ddnu_example}. This history begins in the center
domain and through the course of transitioning through states in the
system arrives at a state that exists in the domain directly to the
left of the center domain. History A must then be communicated from
the center domain to the domain immediately to the left. As with the
particle transport system, we may find that histories in a domain
decomposed Neumann-Ulam method are only communicated a few times
(history B) if at all, or they may be communicated many times (history
C) depending on the discretization and the outcome of the Monte Carlo
game. The amount of communication of histories from domain to domain
will be the primary factor in the parallel scalability of the
algorithm. In the next section, the number of histories communicated in a given
problem will be quantified analytically for a simple model problem.

\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/ddnu_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Domain decomposed Neumann-Ulam example illustrating
      how domain decomposition requires parallel communication of
      histories.} \textit{Each mesh point corresponds to an equation
      in the linear system and the coupling among equations is
      described by the discretization of the problem.}}
  \label{fig:ddnu_example}
\end{figure}

@@

%%---------------------------------------------------------------------------%%
\section{Analytic Performance Framework for Domain-Decomposed Monte Carlo}
\label{sec:analytic_framework}
To date, parallel Neumann-Ulam methods have been limited to full
domain replication with parallelism exploited through individual
histories \citep{alexandrov_efficient_1998} and in this chapter we
will exploit particle transport algorithms to alleviate this. To
accomplish this, we recognize from the literature that stochastic
histories must be transported from domain to domain as the simulation
progresses and they transition to states that are not in the local
domain. Because we have chosen a domain decomposition strategy in a
parallel environment, this means that communication of these histories
must occur between compute nodes owning neighboring pieces of the
global domain. We wish to characterize this communication not only
because communication is in general expensive, but also because these
nearest-neighbor communication sequences, specifically, have poor
algorithmic strong scaling \citep{gropp_high-performance_2001} in much
the same way as a matrix-vector multiply operation. Therefore, we
desire a framework to provide a simple, analytic theory based on the
properties of the linear system that will allow for estimates of the
domain decomposed behavior of the Neumann-Ulam method in terms of the
amount of information that must be communicated.

When solving problems where the linear operator is symmetric, a host
of analytic techniques exist based on the eigenvalue spectrum of the
operator that characterize their behavior in the context of
deterministic linear solvers. Using past work, these techniques are
adapted to the domain decomposed Neumann-Ulam method using the
one-speed, two-dimensional neutron diffusion equation and spatial
discretization presented in
Appendix~\ref{chap:diffusion_problem}. Using the linear system
generated by the discretization of the model problem, we use a
spectral analysis to generate analytic relations for the eigenvalues
of the operator based on system parameters. Using the eigenvalue
spectra, we then build relationships to characterize the transport of
stochastic histories in a decomposed domain and the fraction of
histories that leak from a domain and will therefore have to be
communicated. Finally, we compare these analytic results to numerical
experiments conducted with the model problem.

\subsection{Spectral Analysis}
\label{subsec:spectral_analysis}
The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the eigenvalues of the iteration matrix. We
will compute these eigenvalues by assuming eigenfunctions of the form
\citep{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these eigenfunctions should
give the eigenvalues of the matrix as they exist on the unit circle in
the complex plane.

@@

%%---------------------------------------------------------------------------%%
\section{Domain Decomposed Neumann-Ulam Algorithm}
\label{sec:asynchronous_algorithm}
In the context of radiation transport, in 2009 Brunner and colleagues
provided a fully asynchronous domain decomposed parallel aglorithm as
implemented in production implicit Monte Carlo codes
\citep{brunner_efficient_2009}. In their work they identify two data
sets that are required to be communicated: the sharing of particles
that are transported from one domain to another and therefore from one
processor to another and a global communication that signals if
particle transport has been completed on all processors. Although the
implementation was robust and allowed for scaling to large numbers of
processors, performance issues were still noted with parallel
efficiency improvements needed in both the weak and strong scaling
cases for unbalanced problems. These results led Brunner to conclude
that a combination of domain decomposition and domain replication
could be used to solve some of these issues.

We can take much of what was learned from Brunner's parallel Monte
Carlo method for radiation transport and directly apply it to a
parallel formulation of the Neumann-Ulam method. Direct analogs can be
derived from these works by noting that the primary difference between
solving a linear system with Monte Carlo methods and fixed source
Monte Carlo transport problems is the content of the Markov chains
that are generated. The transitions represented by these chains are
bound by probabilities and weights and are initiated by the sampling
of a source. In the context of transport problems, those transitions
represent events such as particle scattering and absorption with
probabilities that are determined by physical data in the form of
cross sections. For stochastic matrix inversion, those transitions
represent moving between the equations of the linear system (and
therefore the physical domain which they represent) and their
probabilities are defined by the coefficients of those
equations. Ultimately, we tally the contributions to generate
expectation values in the desired states as we progress through the
chains. Therefore, parallel methods for Monte Carlo radiation
transport can be abstracted and we can use those concepts that apply
to matrix inversion methods as an initial means of developing a
parallel Neumann-Ulam-type solver. In this section the fully
asynchronous MSOD Neumann-Ulam algorithm developed for this work is
presented step-by-step in detail.

\subsection{Generating the Transport Domain with Preconditioning}
\label{subsec:domain_generation}

\subsection{Generating the Transport Source with Preconditioning}
\label{subsec:source_generation}

\subsection{Transition Processing Kernel}
\label{subsec:transition_processing}
Processing the transition of a single stochastic history from one
state to another in the system is the fundamental kernel of the
transport procedure. In this kernel, an incoming history has a state
consisting of the current discrete state in the system in which it
resides and its current weight and is processed as given by
Algorithm~\ref{alg:transition_process}.
\begin{algorithm}[h!]
  \caption{Stochastic History Transition Processing Kernel}
  \label{alg:power_iteration}
  \begin{algorithmic}
    \State processTransition( history ):
    \State i = history.state
    \State j = sampleRowCDF( C[i] )
    \State history.state = j
    \State history.weight *= w[i,j]
  \end{algorithmic}
  \label{alg:transition_process}
\end{algorithm}
The core component here is the sampling of the CDF for a particular
state generated by the probablities from the Neumann-Ulam
decomposition. Upon exiting, the kernel has modified the state of the
history with a new state and multiplied the weight with the
corresponding transition weight to form the product in
Eq~(\ref{eq:direct_permutation_weight}).

\subsection{Local Domain Transport Kernel}
\label{subsec:local_domain_transport}

@@

\subsection{Asynchronous Transport Kernel}
\label{subsec:async_transport_kernel}

@@

%%---------------------------------------------------------------------------%%
\section{Multiple-Set Overlapping-Domain Neumann-Ulam Algorithm}
\label{subsec:msod}
In 2010, Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed similar parallel algorithms for particle
transport but a certain amount of overlap between adjacent domains was
used to decrease the number of particles leaving the local domain. In
addition, Wagner utilized a level of replication of the domain such
that the domain was only decomposed on $O(100)$ processors and if
replicated $O(1,000)$ times achieves simulation on $O(100,000)$
processors, thus providing spatial and particle parallelism. Each
collection of processors that constitutes a representation of the
entire domain is referred to as a set, and within a set overlap occurs
among its sub-domains. The original motivation was to decompose the
domain in a way that it remained in a physical cabinet in a large
distributed machine, thus reducing latency costs during
communication. A multiple set scheme is also motivated by the fact
that communication during particle transport only occurs within a set,
limiting communications during the transport procedure to a group of
$O(100)$ processors, a number that was shown to have excellent
parallel efficiencies in Brunner's work and therefore will scale well
in this algorithm. The overlapping domains within each set also
demonstrated reduced communication costs. On each processor, the
source is sampled in the local domain that would exist if no overlap
was used while tallies can be made over the entire overlapping domain.

@@

@@

@@

%%---------------------------------------------------------------------------%%
\section{Small Cluster Scaling Studies}
\label{sec:small_scaling_studies}

@@

\subsection{MSOD Decomposition}
\label{subsec:msod_decomposition}

>>>>

\chapter{Parallel Monte Carlo Synthetic \\ Acceleration Methods}
\label{ch:parallel_methods}
For MCSA to be viable at the production scale for nuclear engineering
applications, scalable parallel implementations of the algorithm are
required. Reviewing the literature, MCSA has yet to be parallelized
and the Neumann-Ulam Monte Carlo method on which it is built has only
been parallelized through history-level parallelism with full domain
replication \citep{alexandrov_efficient_1998}. In order to solve large
linear transport systems with MCSA, a domain-decomposed parallel
strategy is required. In the formulation of a parallel MCSA algorithm,
we recognize that the algorithm occurs in two stages, an outer
iteration performing fixed point iteration and applying the
correction, and an inner Monte Carlo solver that is generating the
correction via the adjoint or forward methods. The parallel aspects of
both these components must be considered. Therefore, we will develop
and implement parallel algorithms for the Neumann-Ulam and MCSA
methods leveraging both the knowledge gained from the general parallel
implementations of Krylov methods reviewed in
Appendix~\ref{ch:linear_problem} and modern parallel strategies for
domain decomposed Monte Carlo as developed by the reactor physics
community.

In this chapter we briefly review particle transport methods for
domain decomposed Monte Carlo. Considering a domain decomposed
Neumann-Ulam method, we derive an analytic framework based on
algebraic quantities from which to estimate its performance when
applied to neutron transport systems. We then devise a parallel
algorithm for the Neumann-Ulam method based on the multiple-set
overlapping-domain decomposition algorithm and a parallel algorithm
for the MCSA iteration that leverages the parallel Monte Carlo
algorithm and general parallel matrix-vector operations. In identical
fashion to the serial computations presented in
Chapter~\ref{ch:spn_equations}, we use parallel solutions for the
$SP_N$ equations for the fuel assembly criticality problem to verify
the correctness of the parallel algorithm and implementation by
comparing the numerical results against production Krylov
methods. Using the verified implementation of the algorithm, we
perform parallel scaling studings to test its performance on a
leadership class machine and compare this performance to the
production Krylov methods.

@@

To motivate this problem, consider the square domain presented in
Figure~\ref{fig:ddmc_example} and on this domain we imagine a Monte
Carlo particle transport problem. If the domain were decomposed into 9
subdomains as shown, each of those subdomains and their associated
data (i.e. cross sections) could be owned by a different parallel
process in the computation. In Figure~\ref{fig:ddmc_example}, the
tracks of 3 particles that are born in the center subdomain are
shown. In this example, particle A is first transported from the
center domain to the domain directly to the left. Before the
scattering event in the new domain may be processed, particle A must
vbe communicated between between the two parallel processes that own
those domains. For any given particle in the transport simulation,
this communication event may hardly occur during the lifetime of the
particle (particle B), or may occur many times (particle C) depending
on the problem parameters and the random path in phase space taken by
the particle. Scalable parallel algorithms for domain decomposed Monte
Carlo are those that handle this domain-to-domain communication of
particles effectively and balance it with the on-process computations
for particle interactions, tallies, response calculations, and other
simulation requirements.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/ddmc_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Domain decomposed Monte Carlo transport example
      illustrating how domain decomposition requires parallel
      communication of particles.}}
  \label{fig:ddmc_example}
\end{figure}

For a domain decomposed Neumann-Ulam method, the situation is very
much the same where instead the transport process is now discrete and
the "physics" of the transport process is the Monte Carlo game with
probabilities and weights described by
Eq~(\ref{eq:neumann_ulam_decomposition}) for the given linear
problem. Figure~\ref{fig:ddnu_example} gives the domain decomposed
Neumann-Ulam analog of particle transport problem in
Figure~\ref{fig:ddmc_example}. Imagine in this problem that we are
solving for the monoenergetic scalar neutron flux as in the model
diffusion problem presented in Appendix~\ref{chap:diffusion_problem}.
In this problem, the solution has only a spatial dependence and
therefore each point in the mesh corresponds to an equation in the
resulting linear system. In Figure~\ref{fig:ddnu_example}, the domain
has now been discretized by this mesh and again decomposed into 9
subdomains, each owning a piece of the mesh and therefore the
corresponding equations in the linear system. As we play the Monte
Carlo game presented in Chapter~\ref{ch:stochastic_methods}, the
discretization and resulting Neumann-Ulam decomposition of the problem
describes how each state in the system is coupled to the other states
in the system through the set of equations that form the linear
system. This coupling is then responsible for the discrete random walk
that each history takes as shown in this example. As discrete states
(or rows in the linear system) that do not exist in the local domain
are reached during the random walk by stochastic histories, the
histories must be communicated to the subdomain that does own the
discrete state in an analogous fashion to the particles in the
previous example.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/ddnu_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Domain decomposed Neumann-Ulam example illustrating
      how domain decomposition requires parallel communication of
      histories.} \textit{Each mesh point corresponds to an equation
      in the linear system and the coupling among equations is
      described by the discretization of the problem.}}
  \label{fig:ddnu_example}
\end{figure}

To further motivate using a parallel algorithm similar to particle
transport methods, again consider history A in
Figure~\ref{fig:ddnu_example}. This history begins in the center
subdomain and through the course of transitioning through states in
the system arrives at a state that exists in the subdomain directly to
the left of the center subdomain. History A must then be communicated
from the center subdomain to the subdomain immediately to the left to
continue the random walk. As with the particle transport system, we
may find that histories in a domain decomposed Neumann-Ulam method are
only communicated a few times (history B) if at all, or they may be
communicated many times (history C) depending on the discretization
and the outcome of the Monte Carlo game. In identical fashion to the
particle transport problem, a domain decomposed Neumann-Ulam method
has the same communcation requirements with all histories in both
examples requiring the same parallel operations. The amount of
communication of histories from domain to domain will be the primary
factor in the parallel scalability of the algorithm. In the next
section, the number of histories communicated in a given problem will
be quantified analytically for a simple model problem.

@@

%%---------------------------------------------------------------------------%%
\section{Analytic Performance Framework for Domain-Decomposed Monte Carlo}
\label{sec:analytic_framework}
To date, parallel Neumann-Ulam methods have been limited to full
domain replication with parallelism exploited through individual
histories \citep{alexandrov_efficient_1998} and in this chapter we
will exploit particle transport algorithms to alleviate this. To
accomplish this, we recognize from the literature that stochastic
histories must be transported from domain to domain as the simulation
progresses and they transition to states that are not in the local
domain. Because we have chosen a domain decomposition strategy in a
parallel environment, this means that communication of these histories
must occur between compute nodes owning neighboring pieces of the
global domain. We wish to characterize this communication not only
because communication is in general expensive, but also because these
nearest-neighbor communication sequences, specifically, have poor
algorithmic strong scaling \citep{gropp_high-performance_2001} in much
the same way as a parallel matrix-vector multiply
operation. Therefore, we desire a framework to provide a simple,
analytic theory based on the properties of the transport system that
will allow for estimates of the domain decomposed behavior of the
Neumann-Ulam method in terms of the amount of information that must be
communicated.

When solving problems where the linear operator is symmetric, a host
of analytic techniques exist based on the eigenvalue spectrum of the
operator that characterize their behavior in the context of
deterministic linear solvers. Using past work, these techniques are
adapted to the domain decomposed Neumann-Ulam method using the
one-speed, two-dimensional neutron diffusion equation and spatial
discretization presented in Appendix~\ref{chap:diffusion_problem} as a
model transport problem. Using the linear system generated by the
discretization of the model problem, we use a spectral analysis to
generate analytic relations for the eigenvalues of the operator based
on system parameters. Using the eigenvalue spectra, we then build
relationships to characterize the transport of stochastic histories in
a decomposed domain and the fraction of histories that leak from a
domain and will therefore have to be communicated. Finally, we compare
these analytic results to numerical experiments conducted with the
model transport problem.

\subsection{Spectral Analysis}
\label{subsec:spectral_analysis}
The convergence of the Neumann series in
Eq~(\ref{eq:adjoint_neumann_series}) approximated by the Monte Carlo
solver is dependent on the eigenvalues of the iteration matrix. We
will compute these eigenvalues by assuming eigenfunctions of the form
\citep{leveque_finite_2007}:
\begin{equation}
  \Phi_{p,q}(x,y) = e^{2 \pi \imath p x} e^{2 \pi \imath q y}\:,
  \label{eq:eigenfunction_form}
\end{equation}
where different combinations of $p$ and $q$ represent the different
eigenmodes of the solution. As these are valid forms of the solution,
then the action of the linear operator on these eigenfunctions will
yield the eigenvalues of the matrix as they exist on the unit circle
in the complex plane.

@@

%%---------------------------------------------------------------------------%%
\section{Domain Decomposed Neumann-Ulam Algorithm}
\label{sec:asynchronous_algorithm}
In the context of radiation transport, in 2009 Brunner and colleagues
provided a fully asynchronous domain decomposed parallel aglorithm as
implemented in production implicit Monte Carlo codes
\citep{brunner_efficient_2009}. We will adapt their algorithm and
directly apply it to a parallel formulation of the Neumann-Ulam
method. Direct analogs can be derived from their works by noting that
the primary difference between solving a linear system with Monte
Carlo methods and fixed source Monte Carlo transport problems is the
content of the Markov chains that are generated. The transitions
represented by these chains are bound by probabilities and weights and
are initiated by the sampling of a discrete source. In the context of
transport problems, those transitions represent events such as
particle scattering and absorption with probabilities that are
determined by physical data in the form of cross sections.

For stochastic matrix inversion, those transitions represent moving
between the equations of the linear system (and therefore the
components of phase space which they represent) and their
probabilities are defined by the coefficients of those
equations. Ultimately, we tally the contributions to generate
expectation values in the desired states as we progress through the
chains. Therefore, parallel methods for Monte Carlo radiation
transport can be abstracted and we can use those concepts that apply
to matrix inversion methods as an initial means of developing a
parallel Neumann-Ulam-type solver.

In this section Brunner and Brantley's fully asynchronous algorithm is
briefly presented along with its application to the Neumann-Ulam
method. Their algorithm is presented in detail in the reference and
was effectively implemented verbatim for this work. In their work they
identify two data sets that are required to be communicated: the
sharing of particles that are transported from one domain to another
and therefore from one processor to another and a global communication
that signals if particle transport has been completed on all
processors. Both of these processes will be addressed at a high level
along with how the parallel Monte Carlo data structures are
constructed.

\subsection{Parallel Transport Domain and Source}
\label{subsec:domain_generation}
To utilize the parallel transport algorithm, we must generate the
parallel transport domain. For the Neumann-Ulam method, the domain
consists of all states in the system that are local, and the
probabilities and weights for all state transitions possible in the
local domain. At the core of this representation is the Neumann-Ulam
decomposition of the linear operator. Given an input decomposition for
the linear operator, by definition the Neumann-Ulam decomposition will
have the same parallel decomposition. In addition, any modification
made to the linear operator through preconditioning or relaxation
parameters will also modify the Neumann-Ulam decomposition. All
possible transitions are described by the local graph of the sparse
input matrix. In addition, any left preconditioning or relaxation
parameters will modify the right hand side of the linear system and
therefore the fixed source in the Monte Carlo calculation. This source
vector will have the same parallel decomposition as the input matrix
and therefore all of the birth states for the histories will exist for
at least a single transition event within the local domain.

@@

\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{ \input{chapters/parallel_mc/domain_to_domain.pdftex_t} }
  \end{center}
  \caption{\textbf{Nearest neighbor history communication sequence.}
    \textit{Each subdomain in the system has a set of nearest
      neighbors determined by the parallel adjacency graph of the
      input matrix. The subdomains are indexed by an integer.}}
  \label{fig:nearest_neighbor_comm}
\end{figure}

\subsection{Ending the Transport Sequence}
\label{subsec:ending_transport}

\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.25}{
      \input{chapters/parallel_mc/binary_comm_tree.pdftex_t} }
    \caption{\textbf{Binary communication tree for coordinating the
        end of a parallel Neumann-Ulam solve.} \textit{Each child
        process reports to its parents how many histories completed
        within its domain. When the root process sums all complete
        histories, it forwards the stop signal to its children which
        in turn forward the message up the tree. The subdomains are
        indexed by an integer.}}
  \end{center}
  \label{fig:binary_comm_tree}
\end{figure}

@@

%%---------------------------------------------------------------------------%%
\section{Multiple-Set Overlapping-Domain Neumann-Ulam Algorithm}
\label{subsec:msod}
Although the implementation presented in the previous section was
observed by Brunner and Brantley to be robust and allowed for scaling
to large numbers of processors, performance issues were still noted
with parallel efficiency improvements needed in both the weak and
strong scaling cases for unbalanced problems. These results led them
to conclude that a combination of domain decomposition and domain
replication could be used to solve some of these issues. In 2010,
Wagner and colleagues developed the \textit{multiple-set
  overlapping-domain} (MSOD) decomposition for parallel Monte Carlo
applications for full-core light water reactor analysis
\citep{wagner_hybrid_2010}. In their work, an extension of Brunner's,
their scheme employed similar parallel algorithms for particle
transport but a certain amount of overlap between adjacent domains was
used to decrease the number of particles leaving the local domain. In
addition, Wagner utilized a level of replication of the domain such
that the domain was only decomposed on $O(100)$ processors and if
replicated $O(1,000)$ times potentially achieves simulation on
$O(100,000)$ processors, thus providing both spatial and particle
parallelism. Each collection of processors that constitutes a
representation of the entire domain is referred to as a set, and
within a set overlap occurs among its sub-domains. The original
motivation was to decompose the domain in a way that it remained in a
physical cabinet in a large distributed machine, thus reducing latency
costs during communication. A multiple set scheme is also motivated by
the fact that communication during particle transport only occurs
within a set, limiting communications during the transport procedure
to a group of $O(100)$ processors, a number that was shown to have
excellent parallel efficiencies in Brunner's work and therefore will
scale well in this algorithm. The overlapping domains within each set
also demonstrated reduced communication costs. On each processor, the
source is sampled in the local domain that would exist if no overlap
was used while tallies can be made over the entire overlapping domain.

@@

\begin{figure}[htpb!]
  \begin{center}
    \scalebox{0.6}{ \input{chapters/parallel_mc/msod_construction.pdftex_t} }
  \end{center}
  \caption{\textbf{MSOD construction.} \textit{The linear system is
      used to construct the Monte Carlo transport domain on the
      primary set. The Monte Carlo data structures are then broadcast
      amongst the blocks to all sets in the system.}}
  \label{fig:msod_construction}
\end{figure}

@@

\begin{figure}[htpb!]
  \begin{center}
    \scalebox{0.6}{ \input{chapters/parallel_mc/msod_tally.pdftex_t} }
  \end{center}
  \caption{\textbf{MSOD tally reduction.} \textit{The tally computed
      independently in each set is reduced across the blocks to the
      primary set. The linear system vector is updated on the primary
      set. In this example the tally vector in the block containing
      upper-left subdomains is combined. Identical reduction
      operations occur for all other blocks simultaneously.}}
  \label{fig:msod_tally_reduction}
\end{figure}

@@

%%---------------------------------------------------------------------------%%
\section{Parallel Performance Metrics}
\label{sec:parallel_performance_metrics}

@@

\subsection{Multiple Set Decomposition}
\label{subsec:msod_decomposition}
